Empirical Analysis of Performance Bottlenecks in Graph Neural Network Training with GPUs

### 2020-08-30 CCPE/Mauscript/sec1-introduction.tex
update

However, what is the real performance bottleneck in the GNN training is still in doubt.
Yan et al. \cite{yan2020_characterizing_gcn} and Zhang et al. \cite{zhang2020_analysis_neugraph} experimentally analyze the architectural characteristics of the GNN \emph{inference}.
They find that the GNN inference is more cache-friendly than the traditional graph analysis tasks(like PageRank) and is suitable for GPU.
They verify the effectiveness of the kernel fusion optimization in reducing the time of inference.
Nevertheless, they only analyze the inference stage, ignoring the effects of the backpropagation during the training.

@@

We implement them with PyG and evaluate their efficiency with six real-world datasets on GPU.
We identify the most time-consuming stage in the GNN training by decomposing the training time per epoch from the layer level to the operator level.
We also analyze the memory usage during the training to discover the main factor that limits the data scalability of GNN training on GPU. 
Finally, we evaluate whether or not the sampling techniques affect the performance bottleneck. 
The key findings and insights are summarized below.

>>>>

However, what is the real performance bottleneck in the GNN training is still in doubt.
Yan et al. \cite{yan2020_characterizing_gcn} and Zhang et al. \cite{zhang2020_analysis_neugraph} experimentally analyze the architectural characteristics of the GNN \emph{inference}.
They find that the GNN inference is more cache-friendly than the traditional graph analysis tasks(like PageRank) and is suitable for GPUs.
They verify the effectiveness of the kernel fusion optimization in reducing the time of inference.
Nevertheless, they only analyze the inference stage, ignoring the effects of the backpropagation during the training.

@@

We implement them with PyG and evaluate their efficiency with six real-world datasets on a GPU.
We identify the most time-consuming stage in the GNN training by decomposing the training time per epoch from the layer level to the operator level.
We also analyze the memory usage during the training to discover the main factor that limits the data scalability of GNN training on GPUs. 
Finally, we evaluate whether or not the sampling techniques affect the performance bottleneck. 
The key findings and insights are summarized below.

### 2020-08-30 CCPE/Mauscript/sec4-experiments.tex
update

The functions $\phi$, $\Sigma$ and $\gamma$ in the vertex and edge calculation are made up of a series of basic operators implemented on GPU, like the matrix multiplication \texttt{mm}, the elementwise multiplication \texttt{mul} and the index-based selection \texttt{index\_select}.
\figurename~\ref{fig:exp_top_basic_ops} shows the five most time-consuming basic operators in each GNN, averaged over all the real-world graphs in \tablename~\ref{tab:dataset_overview}.

@@

In general, the most time-consuming operator in the four GNNs, is still the matrix multiplication \texttt{mm} and the elementwise multiplication \texttt{mul}, \emph{making GNN training suitable for GPU}.
Although the aggregation step in the edge calculation stage is relatively simple (like sum and mean), the related operators \texttt{scatter\_add} and \texttt{gather} still consume a certain amount of the time.
The operators have to synchronize between hardware threads to avoid updating the same aggregated vector at the same time. They also conduct non-regular memory access with the access pattern determined by the edge set dynamically.
For GPUs, they are less efficient than \texttt{mm}.
The index-based selection \texttt{index\_select} operator from the collection step consume around 10\% of the time in all GNNs.
Though GPUs have high on-chip memory bandwidth, improving the efficiency of \texttt{scatter\_add}/\texttt{gather}/\texttt{index\_select} can benefit the training of all kinds of GNNs.

\paragraph{Summary}
The GNN training is suitable for GPU.
\textbf{The edge calculation stage is the main performance bottleneck in most cases}, except for training GNNs with high vertex calculation complexity on low-average-degree graphs.
The performance bottleneck in the edge calculation stage depends on the time complexity of the messaging function $\phi$.
\begin{itemize}
    \item If the time complexity of $\phi$ is \textbf{high}, the \textbf{efficiency of $\phi$} limits the performance. Reducing its computation cost (via optimizing its implementation or modifying the algorithm) can significantly reduce the training time.
    \item If the time complexity of $\phi$ is \textbf{low}, the \textbf{collection step} and the \textbf{aggregation step} limit the performance. The collection step involves lots of data movement. The aggregation step suffers from data synchronization and non-regular data access. Optimizing their implementations can significantly reduce the training time.
\end{itemize}

@@

During the GNN training, all data (including datasets and intermediate results) are stored in the on-chip memory of GPU.
Compared with the main memory on the host side, the capacity of the GPU memory is very limited.
\emph{The GPU memory capacity limits the scales of the graphs that a GPU can train GNNs on}.
For example, GaAN is unable to train on \texttt{cph} dataset due to the out of memory exception.

@@

In the implementation in PyG, the GNN model and the dataset resides on the GPU side.
To process each epoch, PyG samples the original dataset in the main memory and generates several batches.
Each batch is a small subgraph of the dataset.
To train on each batch, PyG sends the sampled subgraph to GPU, calculates the gradients on the subgraph, and updates the model parameters directly on GPU.
With the sampling techniques, the model parameters are updated by a stochastic gradient descent optimizer.
PyG conducts the evaluation phase every several epochs (either on the CPU side or the GPU side) to determine whether to stop the training.
In this section, the experiments focus on the training phase of each batch.

@@

To find out the performance bottleneck with the sampling techniques, we decompose the training time per batch into three phases: \emph{sampling} on CPU, \emph{transferring} sampled subgraphs from CPU to GPU and \emph{training} with the subgraphs on GPU.
\figurename~\ref{fig:exp_sampling_batch_train_time} shows the time breakdown of the four GNNs under different relative batch sizes.
For the neighbor sampler, the sampling technique reduces the training time per batch only when the batch size is very small.
When the batch becomes bigger, the sampling and the data transferring phases introduce noticeable overheads, making the training time exceed the full-batch training.
For the clustering sampler, the sampled subgraph is smaller than the neighbor sampler under the same relative batch size.
The reduction in the training time is more obvious than the neighbor sampler.
However, the overheads increase quickly as the relative batch size increase.
The training time under the 25\% relative batch size already exceede the time of full-batch training.
The experimental results indicate that the current implementation of the sampling techniques in PyG is inefficient.
When the batch size is slightly big, more than 50\% of the time has been spent on sampling and data transferring.
\emph{The sampling techniques are only efficient under small batch sizes.}

>>>>

The functions $\phi$, $\Sigma$ and $\gamma$ in the vertex and edge calculation are made up of a series of basic operators implemented on the GPU, like the matrix multiplication \texttt{mm}, the elementwise multiplication \texttt{mul} and the index-based selection \texttt{index\_select}.
\figurename~\ref{fig:exp_top_basic_ops} shows the five most time-consuming basic operators in each GNN, averaged over all the real-world graphs in \tablename~\ref{tab:dataset_overview}.

@@

In general, the most time-consuming operator in the four GNNs, is still the matrix multiplication \texttt{mm} and the elementwise multiplication \texttt{mul}, \emph{making GNN training suitable for GPUs}.
Although the aggregation step in the edge calculation stage is relatively simple (like sum and mean), the related operators \texttt{scatter\_add} and \texttt{gather} still consume a certain amount of the time.
The operators have to synchronize between hardware threads to avoid updating the same aggregated vector at the same time. They also conduct non-regular memory access with the access pattern determined by the edge set dynamically.
For GPUs, they are less efficient than \texttt{mm}.
The index-based selection \texttt{index\_select} operator from the collection step consume around 10\% of the time in all GNNs.
Though GPUs have high on-chip memory bandwidth, improving the efficiency of \texttt{scatter\_add}/\texttt{gather}/\texttt{index\_select} can benefit the training of all kinds of GNNs.

\paragraph{Summary}
The GNN training is suitable for GPUs.
\textbf{The edge calculation stage is the main performance bottleneck in most cases}, except for training GNNs with high vertex calculation complexity on low-average-degree graphs.
The performance bottleneck in the edge calculation stage depends on the time complexity of the messaging function $\phi$.
\begin{itemize}
    \item If the time complexity of $\phi$ is \textbf{high}, the \textbf{efficiency of $\phi$} limits the performance. Reducing its computation cost (via optimizing its implementation or modifying the algorithm) can significantly reduce the training time.
    \item If the time complexity of $\phi$ is \textbf{low}, the \textbf{collection step} and the \textbf{aggregation step} limit the performance. The collection step involves lots of data movement. The aggregation step suffers from data synchronization and non-regular data access. Optimizing their implementations can significantly reduce the training time.
\end{itemize}

@@

During the GNN training, all data (including datasets and intermediate results) are stored in the on-chip memory of the GPU.
Compared with the main memory on the host side, the capacity of the GPU memory is very limited.
\emph{The GPU memory capacity limits the scales of the graphs that a GPU can train GNNs on}.
For example, GaAN is unable to train on \texttt{cph} dataset due to the out of memory exception.

@@

In the implementation in PyG, the GNN model and the dataset resides on the GPU side.
To process each epoch, PyG samples the original dataset in the main memory and generates several batches.
Each batch is a small subgraph of the dataset.
To train on each batch, PyG sends the sampled subgraph to the GPU, calculates the gradients on the subgraph, and updates the model parameters directly on GPU.
With the sampling techniques, the model parameters are updated by a stochastic gradient descent optimizer.
PyG conducts the evaluation phase every several epochs (either on the CPU side or the GPU side) to determine whether to stop the training.
In this section, the experiments focus on the training phase of each batch.

@@

To find out the performance bottleneck with the sampling techniques, we decompose the training time per batch into three phases: \emph{sampling} on the CPU, \emph{transferring} sampled subgraphs from the CPU to the GPU and \emph{training} with the subgraphs on GPU.
\figurename~\ref{fig:exp_sampling_batch_train_time} shows the time breakdown of the four GNNs under different relative batch sizes.
For the neighbor sampler, the sampling technique reduces the training time per batch only when the batch size is very small.
When the batch becomes bigger, the sampling and the data transferring phases introduce noticeable overheads, making the training time exceed the full-batch training.
For the clustering sampler, the sampled subgraph is smaller than the neighbor sampler under the same relative batch size.
The reduction in the training time is more obvious than the neighbor sampler.
However, the overheads increase quickly as the relative batch size increase.
The training time under the 25\% relative batch size already exceede the time of full-batch training.
The experimental results indicate that the current implementation of the sampling techniques in PyG is inefficient.
When the batch size is slightly big, more than 50\% of the time has been spent on sampling and data transferring.
\emph{The sampling techniques are only efficient under small batch sizes.}

### 2020-08-30 CCPE/Mauscript/sec6-related-work.tex
update

From the efficiency aspect, Yan et al. \cite{yan2020_characterizing_gcn} compare the performance characteristics of graph convolutional networks, typical graph processing kernels(like PageRank), and the MLP-based neural networks on GPU.
They provide optimization guidelines for both the software and the hardware developers.
Zhang et al. \cite{zhang2020_analysis_neugraph} analyze the architectural characteristics of the GNN inference on GPU under SAGA-NN \cite{ma2019_neugraph} model.
They find that the GNN inference has no fixed performance bottleneck and all components deserve to optimize.
These two efforts focus on the \emph{inference} phase of GNNs and they investigate the potential optimizations mainly from an architectural view.
In this work, our target is to find out the performance bottleneck in the training phase from a system view.
We consider the performance bottleneck in both time and memory usage.
We also evaluate the effects of the sampling techniques.
Our work and \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph} form a complementary study on the efficiency issue of GNNs.

>>>>

From the efficiency aspect, Yan et al. \cite{yan2020_characterizing_gcn} compare the performance characteristics of graph convolutional networks, typical graph processing kernels(like PageRank), and the MLP-based neural networks on GPUs.
They provide optimization guidelines for both the software and the hardware developers.
Zhang et al. \cite{zhang2020_analysis_neugraph} analyze the architectural characteristics of the GNN inference on GPUs under SAGA-NN \cite{ma2019_neugraph} model.
They find that the GNN inference has no fixed performance bottleneck and all components deserve to optimize.
These two efforts focus on the \emph{inference} phase of GNNs and they investigate the potential optimizations mainly from an architectural view.
In this work, our target is to find out the performance bottleneck in the training phase from a system view.
We consider the performance bottleneck in both time and memory usage.
We also evaluate the effects of the sampling techniques.
Our work and \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph} form a complementary study on the efficiency issue of GNNs.
